---
title: "Ecological Genomics Homework 2"
author: "Aidan Jones"
date: "2025-11-01"
output: html_document
---

## Ecological Genomics Homework 2 - Transcriptomics

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/users/a/j/ajones76/projects/eco_genomics_2025/transcriptomics/mydata")
```

### Import required libraries and dependencies

```{r}
library(DESeq2)
library(topGO)
library(GO.db)
library(dplyr)
library(ggplot2)
library(tidyr)
library(purrr)
```

## Import counts matrix and sample distibution table

```{r}
# Import the counts matrix
countsTable <- read.table("counts_matrix.txt", header=TRUE, row.names=1)
countsTableRound <- round(countsTable) # Round counts to be compatible with DESeq2

# Import the sample description table
conds <- read.delim("metadata.txt", header=TRUE, stringsAsFactors = TRUE, row.names=1)

# Corrects the column names to match between the metadata and the counts matrix
colnames(countsTableRound) <- substr(colnames(countsTableRound), start = 1, stop = 4)
```

### Run DESeq2 to test for differential gene expression

```{r}
# Create a DESeq object and define the experimental design here with the tilda
dds <- DESeqDataSetFromMatrix(countData = countsTableRound, colData=conds, 
                              design= ~ generation + line + generation:line)

# Filter out genes with too few reads - remove all genes with counts < 15 in more than 75% of 24 samples, so 18 (suggested by WGCNA on RNAseq FAQ)
dds <- dds[rowSums(counts(dds) >= 15) >= 18,]

# Relevel factors by defining reference for line and generation
dds$line <- relevel(dds$line, ref = "control")
dds$generation <- relevel(dds$generation, ref = "G1")

# Run the DESeq model to test for differential gene expression
dds <- DESeq(dds)

# List the results you've generated
resultsNames(dds)
# Copy the results names 
```

### Set up contrasts between generations

```{r}
# Extract only generation-specific treatment effects with numeric contrast vectors

# Control G1 vs Treatment G1 (Plastic Response)
res_G1vG1 <- results(dds, name = "line_treatment_vs_control", alpha = 0.05)
res_G1vG1 <- res_G1vG1[order(res_G1vG1$padj), ]
res_G1vG1 <- res_G1vG1[!is.na(res_G1vG1$padj), ]
degs_G1vG1 <- row.names(res_G1vG1[res_G1vG1$padj < 0.05, ])

# G1 vs G2 (Intergenerational Effect)
res_G1vG2 <- results(
  dds,
  list(
    c("generation_G2_vs_G1", "generationG2.linetreatment")),
  alpha = 0.05)
res_G1vG2 <- res_G1vG2[order(res_G1vG2$padj), ]
res_G1vG2 <- res_G1vG2[!is.na(res_G1vG2$padj), ]
degs_G1vG2 <- row.names(res_G1vG2[res_G1vG2$padj < 0.05, ])

# Treatment G2 vs Treatment G3 (Transgenerational Effect)
res_G2vG3 <- results(
  dds,
  list(
    c("generation_G3_vs_G1", "generationG3.linetreatment"),
    c("generation_G2_vs_G1", "generationG2.linetreatment")), 
  alpha = 0.05)
res_G2vG3 <- res_G2vG3[order(res_G2vG3$padj), ]
res_G2vG3 <- res_G2vG3[!is.na(res_G2vG3$padj), ]
degs_G2vG3 <- row.names(res_G2vG3[res_G2vG3$padj < 0.05, ])

# G3 vs G4 (Transgenerational Effect)
res_G3vG4 <- results(
  dds,
  list(
    c("generation_G4_vs_G1", "generationG4.linetreatment"),
    c("generation_G3_vs_G1", "generationG3.linetreatment")), 
  alpha = 0.05)
res_G3vG4 <- res_G3vG4[order(res_G3vG4$padj), ]
res_G3vG4 <- res_G3vG4[!is.na(res_G3vG4$padj), ]
degs_G3vG4 <- row.names(res_G3vG4[res_G3vG4$padj < 0.05, ])

# Function to extract the number of up-regulated and down-regulated genes in the constrasts as well as the mean log2 fold changes for each group
get_mean_lfc <- function(res_obj, contrast_name) {
  res_df <- as.data.frame(res_obj)
  res_df <- res_df[!is.na(res_df$padj), ]
  
  # Identify significantly up-regulated and down-regulated genes
  up_genes <- res_df %>% filter(padj < 0.05, log2FoldChange > 0)
  down_genes <- res_df %>% filter(padj < 0.05, log2FoldChange < 0)
  
  # Compute means
  mean_up <- mean(up_genes$log2FoldChange, na.rm = TRUE)
  mean_down <- mean(down_genes$log2FoldChange, na.rm = TRUE)
  
  # Return caclculations as a named vector
  data.frame(
    Contrast = contrast_name,
    Upregulated_Mean_LFC = mean_up,
    Downregulated_Mean_LFC = mean_down,
    Num_Up = nrow(up_genes),
    Num_Down = nrow(down_genes)
  )
}

# Apply function to all contrasts
mean_lfc_summary <- map2_dfr(contrast_list, names(contrast_list),
                               ~get_mean_lfc(.x, .y))

# Display results summary
mean_lfc_summary

```

## Prepare contrasts and GO terms for TopGO

```{r}
# Define contrasts
contrast_list <- list(
  "Control G1 vs. Treatment G1" = res_G1vG1,
  "Treatment G1 vs. Treatment G2" = res_G1vG2,
  "Treatment G2 vs. Treatment G3" = res_G2vG3,
  "Treatment G3 vs. Treatment G4" = res_G3vG4
)

# Import GO terms file
tonsa_go <- read.delim("Genes_GO_terms_output.tsv", header = TRUE)
tonsa_go <- as.data.frame(tonsa_go)
tonsa_go <- tonsa_go[-c(1,2), ]
tonsa_go <- na.omit(tonsa_go)

# Create gene2GO list
gene2GO <- tonsa_go %>%
  filter(!is.na(GO) & GO != "") %>%
  group_by(geneID) %>%
  summarise(GO = list(unique(GO)), .groups = "drop") %>%
  with(setNames(GO, geneID))

# Split semicolon-separated GO terms
gene2GO <- lapply(gene2GO, function(x) unlist(strsplit(x, ";")))
```

## Generate volvano plots to Visualize abundance and directionality of differential genes expression

```{r}
# Function to make a volcano plot from a DESeq2 results object
make_volcano_plot <- function(res_obj, contrast_name, lfc_thresh = 1, padj_thresh = 0.05) {
  
  # Convert to data frame and remove NAs
  res_df <- as.data.frame(res_obj)
  res_df$geneID <- rownames(res_df)
  res_df <- res_df %>% filter(!is.na(padj))
  
  # Add significance column for color mapping
  res_df$significance <- "Not Sig"
  res_df$significance[res_df$padj < 0.05 & res_df$log2FoldChange > 0] <- "Up"
  res_df$significance[res_df$padj < 0.05 & res_df$log2FoldChange < 0] <- "Down"
  
  # Create volcano plot
  p <- ggplot(res_df, aes(x = log2FoldChange, y = -log10(padj), color = significance)) +
    geom_point(alpha = 0.7, size = 2) +
    scale_color_manual(values = c("Up" = "red", "Down" = "blue", "Not Sig" = "grey")) +
    geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black") +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
    labs(
      title = paste0("Volcano Plot: ", contrast_name),
      x = expression(log[2]~Fold~Change),
      y = expression(-log[10]~Adjusted~p~value)
    ) +
    theme_bw(base_size = 16) +
    theme(legend.title = element_blank(),
          plot.title = element_text(hjust = 0.5))
  
  return(p)
}

# Display all plots
volcano_plots <- map2(contrast_list, names(contrast_list),
                      ~make_volcano_plot(.x, .y))
volcano_plots
```

## Run TopGO for each contrast

```{r}
# Define TopGO function
run_topGO_plot <- function(res_obj, contrast_name, gene2GO, outfile_prefix="topGO") {
  
  # Convert results to data frame and add geneID
  res_df <- as.data.frame(res_obj)
  res_df$geneID <- rownames(res_df)
  
  # Keep only significant DEGs with GO annotations
  res_GO_sig <- res_df %>%
    filter(padj < 0.05 & abs(log2FoldChange) > 1 & geneID %in% names(gene2GO))
  
  # Create geneList for TopGO: 1 = DEG, 0 = not DEG
  all_genes <- names(gene2GO)
  geneList <- setNames(as.integer(all_genes %in% res_GO_sig$geneID), all_genes)
  
  # Build TopGO object
  GOdata <- new("topGOdata",
                description = paste0("DEGs for ", contrast_name),
                ontology = "BP",
                allGenes = geneList,
                geneSel = function(x) x == 1,
                nodeSize = 5,
                annot = annFUN.gene2GO,
                gene2GO = gene2GO)
  
  # Run enrichment test
  result <- runTest(GOdata, algorithm = "parentChild", statistic = "fisher")
  
  # Convert results to table
  res_table <- GenTable(GOdata, classicFisher = result,
                        topNodes = length(usedGO(GOdata))) %>%
    mutate(classicFisher = as.numeric(classicFisher),
           Annotated = as.numeric(Annotated),
           Significant = as.numeric(Significant))
  
  # Filter GO terms
  filtered_GO_results <- res_table %>%
    filter(Annotated >= 5 & Annotated < 500) %>%
    arrange(classicFisher)
  
  # Prepare REVIGO input
  revigo_input <- filtered_GO_results %>%
    select(GO.ID, classicFisher) %>%
    filter(classicFisher < 0.05)
  colnames(revigo_input) <- c("GO.ID", "pValue")
  
  # Write REVIGO file
  outfile <- paste0(outfile_prefix, "_", contrast_name, "_REVIGO.txt")
  write.table(revigo_input, file = outfile, sep = "\t", row.names = FALSE, quote = FALSE)
  
  # Generate TopGO plot
  p <- filtered_GO_results %>%
    slice_head(n = 30) %>%
    ggplot(aes(x = -log10(classicFisher),
               y = reorder(Term, -log10(classicFisher)),
               size = Significant,
               color = -log10(classicFisher))) +
    geom_point(alpha = 0.8) +
    scale_color_gradientn(colors = c("pink", "orchid", "purple", "purple4"),
                          name = "-log10(p-value)") +
    scale_size_continuous(name = "# Significant Genes") +
    xlab("-log10-classicFisher") +
    ylab("Biological Function") +
    ggtitle(paste0("TopGO: ", contrast_name)) +
    theme_minimal() +
    theme(plot.margin = margin(10, 100, 10, 10),
          axis.text.y = element_text(size = 10, hjust = 1)) +
    coord_cartesian(clip = "off")
  
  return(p)
}

# Run TopGO for all contrasts
topGO_plots <- map2(contrast_list, names(contrast_list),
                    ~run_topGO_plot(.x, .y, gene2GO))
topGO_plots
```