---
title: "Ecological Genomics Homework 2"
author: "Aidan Jones"
date: "2025-11-01"
output: html_document
---

## Ecological Genomics Homework 2 - Transcriptomics

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/users/a/j/ajones76/projects/eco_genomics_2025/transcriptomics/mydata")
```

### Import required libraries and dependencies

```{r}
library(DESeq2)
library(topGO)
library(GO.db)
library(dplyr)
library(ggplot2)
library(tidyr)
library(purrr)
```

## Import counts matrix and sample distibution table

```{r}
# Import the counts matrix
countsTable <- read.table("counts_matrix.txt", header=TRUE, row.names=1)
countsTableRound <- round(countsTable) # Round counts to be compatible with DESeq2

# Import the sample description table
conds <- read.delim("metadata.txt", header=TRUE, stringsAsFactors = TRUE, row.names=1)

# Corrects the column names to match between the metadata and the counts matrix
colnames(countsTableRound) <- substr(colnames(countsTableRound), start = 1, stop = 4)
```

### Run DESeq2 to test for differential gene expression

```{r}
# Create a DESeq object and define the experimental design here with the tilda
dds <- DESeqDataSetFromMatrix(countData = countsTableRound, colData=conds, 
                              design= ~ generation + line + generation:line)

# Filter out genes with too few reads - remove all genes with counts < 15 in more than 75% of 24 samples, so 18 (suggested by WGCNA on RNAseq FAQ)
dds <- dds[rowSums(counts(dds) >= 15) >= 18,]

# Relevel factors by defining reference for line and generation
dds$line <- relevel(dds$line, ref = "control")
dds$generation <- relevel(dds$generation, ref = "G1")

# Run the DESeq model to test for differential gene expression
dds <- DESeq(dds)

# List the results you've generated
resultsNames(dds)
# Copy the results names 
```

### Set up contrasts between generations

```{r}
# Extract only generation-specific treatment effects with numeric contrast vectors

# Control G1 vs Treatment G1 (Plastic Response)
res_G1vG1 <- results(dds, name = "line_treatment_vs_control", alpha = 0.05)
res_G1vG1 <- res_G1vG1[order(res_G1vG1$padj), ]
res_G1vG1 <- res_G1vG1[!is.na(res_G1vG1$padj), ]
degs_G1vG1 <- row.names(res_G1vG1[res_G1vG1$padj < 0.05, ])

# G1 vs G2 (Intergenerational Effect)
res_G1vG2 <- results(
  dds,
  list(
    c("generation_G2_vs_G1", "generationG2.linetreatment")),
  alpha = 0.05)
res_G1vG2 <- res_G1vG2[order(res_G1vG2$padj), ]
res_G1vG2 <- res_G1vG2[!is.na(res_G1vG2$padj), ]
degs_G1vG2 <- row.names(res_G1vG2[res_G1vG2$padj < 0.05, ])

# Treatment G2 vs Treatment G3 (Transgenerational Effect)
res_G2vG3 <- results(
  dds,
  list(
    c("generation_G3_vs_G1", "generationG3.linetreatment"),
    c("generation_G2_vs_G1", "generationG2.linetreatment")), 
  alpha = 0.05)
res_G2vG3 <- res_G2vG3[order(res_G2vG3$padj), ]
res_G2vG3 <- res_G2vG3[!is.na(res_G2vG3$padj), ]
degs_G2vG3 <- row.names(res_G2vG3[res_G2vG3$padj < 0.05, ])

# G3 vs G4 (Transgenerational Effect)
res_G3vG4 <- results(
  dds,
  list(
    c("generation_G4_vs_G1", "generationG4.linetreatment"),
    c("generation_G3_vs_G1", "generationG3.linetreatment")), 
  alpha = 0.05)
res_G3vG4 <- res_G3vG4[order(res_G3vG4$padj), ]
res_G3vG4 <- res_G3vG4[!is.na(res_G3vG4$padj), ]
degs_G3vG4 <- row.names(res_G3vG4[res_G3vG4$padj < 0.05, ])

# Function to extract the number of up-regulated and down-regulated genes in the constrasts as well as the mean log2 fold changes for each group
get_lfc_summary <- function(res_obj, contrast_name) {
  res_df <- as.data.frame(res_obj)
  res_df <- res_df[!is.na(res_df$padj), ]
  
  # Identify significantly up-regulated and down-regulated genes
  up_genes <- res_df %>% filter(padj < 0.05, log2FoldChange > 0)
  down_genes <- res_df %>% filter(padj < 0.05, log2FoldChange < 0)
  
  # Compute means and medians
  mean_up <- mean(up_genes$log2FoldChange, na.rm = TRUE)
  median_up <- median(up_genes$log2FoldChange, na.rm = TRUE)
  mean_down <- mean(down_genes$log2FoldChange, na.rm = TRUE)
  median_down <- median(down_genes$log2FoldChange, na.rm = TRUE)
  
  # Return summary statistics
  data.frame(
    Contrast = contrast_name,
    Num_Up = nrow(up_genes),
    Num_Down = nrow(down_genes),
    Upregulated_Mean_LFC = mean_up,
    Upregulated_Median_LFC = median_up,
    Downregulated_Mean_LFC = mean_down,
    Downregulated_Median_LFC = median_down
  )
}

# Apply function to all contrasts
lfc_summary <- map2_dfr(contrast_list, names(contrast_list),
                        ~get_lfc_summary(.x, .y))

# Display results summary
lfc_summary
```

## Prepare contrasts and GO terms for TopGO

```{r}
# Define contrasts
contrast_list <- list(
  "Control G1 vs. Treatment G1" = res_G1vG1,
  "Treatment G1 vs. Treatment G2" = res_G1vG2,
  "Treatment G2 vs. Treatment G3" = res_G2vG3,
  "Treatment G3 vs. Treatment G4" = res_G3vG4
)

# Import GO terms file
tonsa_go <- read.delim("Genes_GO_terms_output.tsv", header = TRUE)
tonsa_go <- as.data.frame(tonsa_go)
tonsa_go <- tonsa_go[-c(1,2), ]
tonsa_go <- na.omit(tonsa_go)

# Create gene2GO list
gene2GO <- tonsa_go %>%
  filter(!is.na(GO) & GO != "") %>%
  group_by(geneID) %>%
  summarise(GO = list(unique(GO)), .groups = "drop") %>%
  with(setNames(GO, geneID))

# Split semicolon-separated GO terms
gene2GO <- lapply(gene2GO, function(x) unlist(strsplit(x, ";")))
```

## Generate volvano plots to visualize abundance and directionality of differential genes expression

```{r}
# Define volcano plot function
make_volcano_plot <- function(res_obj, contrast_name, lfc_thresh = 1, padj_thresh = 0.05) {
  
  # Convert to data frame and remove NAs
  res_df <- as.data.frame(res_obj)
  res_df$geneID <- rownames(res_df)
  res_df <- res_df %>% filter(!is.na(padj))
  
  # Add significance column for color mapping
  res_df$significance <- "Not Sig"
  res_df$significance[res_df$padj < 0.05 & res_df$log2FoldChange > 0] <- "Up"
  res_df$significance[res_df$padj < 0.05 & res_df$log2FoldChange < 0] <- "Down"
  
  # Create volcano plot
  p <- ggplot(res_df, aes(x = log2FoldChange, y = -log10(padj), color = significance)) +
    geom_point(alpha = 0.7, size = 2) +
    scale_color_manual(values = c("Up" = "red", "Down" = "blue", "Not Sig" = "grey")) +
    geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black") +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
    labs(
      title = paste0("Volcano Plot: ", contrast_name),
      x = expression(log[2]~Fold~Change),
      y = expression(-log[10]~Adjusted~p~value)
    ) +
    theme_bw(base_size = 16) +
    theme(legend.title = element_blank(),
          plot.title = element_text(hjust = 0.5))
  
  return(p)
}

# Generate and save volcano plots for all contrasts
walk2(contrast_list, names(contrast_list), function(res_obj, contrast_name) {
  
  # Create the plot
  p <- make_volcano_plot(res_obj, contrast_name)
  
  # Save it as PNG
  ggsave(
    filename = paste0("Volcano_", gsub(" ", "_", contrast_name), ".png"),
    plot = p, width = 8, height = 6, dpi = 300)
})
```

## Run TopGO for each contrast while subsetting up-regulated and down-regulated genes

```{r}
# Define TopGO function
run_topGO_plot <- function(res_obj, contrast_name, gene2GO, outfile_prefix="topGO") {
  
  # Convert results to data frame and add geneID
  res_df <- as.data.frame(res_obj)
  res_df$geneID <- rownames(res_df)
  
  # Split DEGs into up-regulated and down-regulated subsets
  res_up <- res_df %>%
    filter(padj < 0.05 & log2FoldChange > 1 & geneID %in% names(gene2GO))
  
  res_down <- res_df %>%
    filter(padj < 0.05 & log2FoldChange < -1 & geneID %in% names(gene2GO))
  
  # Define internal TopGO function
  run_topGO_sub <- function(res_sig, direction_label, color_scheme) {
    if (nrow(res_sig) == 0) return(NULL)
    
    # Create geneList for TopGO: 1 = DEG, 0 = not DEG
    all_genes <- names(gene2GO)
    geneList <- setNames(as.integer(all_genes %in% res_sig$geneID), all_genes)
    
    # Build TopGO object
    GOdata <- new("topGOdata",
                  description = paste0("DEGs for ", contrast_name, " (", direction_label, ")"),
                  ontology = "BP",
                  allGenes = geneList,
                  geneSel = function(x) x == 1,
                  nodeSize = 5,
                  annot = annFUN.gene2GO,
                  gene2GO = gene2GO)
    
    # Run enrichment test
    result <- runTest(GOdata, algorithm = "parentChild", statistic = "fisher")
    
    # Convert results to table
    res_table <- GenTable(GOdata, classicFisher = result,
                          topNodes = length(usedGO(GOdata))) %>%
      mutate(classicFisher = as.numeric(classicFisher),
             Annotated = as.numeric(Annotated),
             Significant = as.numeric(Significant))
    
    # Filter GO terms
    filtered_GO_results <- res_table %>%
      filter(Annotated >= 5 & Annotated < 500) %>%
      arrange(classicFisher)
    
    # Prepare REVIGO input
    revigo_input <- filtered_GO_results %>%
      select(GO.ID, classicFisher) %>%
      filter(classicFisher < 0.05)
    colnames(revigo_input) <- c("GO.ID", "pValue")
    
    # Write REVIGO file
    outfile <- paste0(outfile_prefix, "_", contrast_name, "_", direction_label, "_REVIGO.txt")
    write.table(revigo_input, file = outfile, sep = "\t", row.names = FALSE, quote = FALSE)
    
    # Generate TopGO plot
    p <- filtered_GO_results %>%
      slice_head(n = 30) %>%
      ggplot(aes(x = -log10(classicFisher),
                 y = reorder(Term, -log10(classicFisher)),
                 size = Significant,
                 color = -log10(classicFisher))) +
      geom_point(alpha = 0.8) +
      scale_color_gradientn(colors = color_scheme,
                            name = "-log10(p-value)") +
      scale_size_continuous(name = "# Significant Genes") +
      xlab("-log10-classicFisher") +
      ylab("Biological Function") +
      ggtitle(paste0("TopGO: ", contrast_name, " (", direction_label, ")")) +
      theme_minimal() +
      theme(
        plot.title = element_text(hjust = 0, size = 16, face = "bold"),
        plot.title.position = "plot",          
        plot.margin = margin(10, 40, 10, 10),
        axis.text.y = element_text(size = 10, hjust = 1)
        ) +
      coord_cartesian(clip = "off")

    return(p)
  }
  
  # Run TopGO for up-regulated and down-regulated subsets
  p_up <- run_topGO_sub(res_up, "Upregulated", c("pink", "orchid", "purple", "purple4"))
  p_down <- run_topGO_sub(res_down, "Downregulated", c("skyblue", "lightblue3", "blue3", "navy"))
  
  return(list(up = p_up, down = p_down))
}

# Run TopGO for all contrasts and save plots
walk2(contrast_list, names(contrast_list), function(res_obj, contrast_name) {
  
  # Run TopGO
  plots <- run_topGO_plot(res_obj, contrast_name, gene2GO)
  
  # Save plots of up-regulated and down-regulated genes
  ggsave(paste0("TopGO_", contrast_name, "_upregulated.png"),
         plot = plots$up, width = 8, height = 6, dpi = 300)
  
  ggsave(paste0("TopGO_", contrast_name, "_downregulated.png"),
         plot = plots$down, width = 8, height = 6, dpi = 300)
})
```
